---
title: "02_odds_ratio_prep.qmd"
author: "Maiko Hata"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(readxl)
library(scales)
library(reactable)
library(dplyr)
library(epitools)
library(rio)
library(corrplot)
library(pwr)
library(epitools)
library(shiny)
library(plotly)
library(readr)

```

✅ Step 1. Make paths stable (so you don’t get this error again)

```{r}
setwd("/Users/maiko/Desktop/capstone")
getwd()
```

✅ Step 1.1 — Read your RDS and do a quick peek

```{r}
setwd("..")
getwd()
list.files("data/analysis")
```

```{r}
df <- readRDS("/Users/maiko/Desktop/capstone/data/analysis/osep_exits_all_years_all_race_analysis.rds")

dim(df)
names(df)

```

✅ Step 1.4 — Standardize column names + confirm the required fields exist

```{r}
# standardize names to lowercase for consistency
names(df) <- tolower(names(df))

# define exit category columns
exit_cats <- c(
  "dismissed",
  "moved_out",
  "not_determined",
  "not_eligible",
  "part_b_eligible",
  "withdrawn"
)

# required columns (keep race_ethnicity)
required_cols <- c("state", "race_ethnicity", "total_exits", exit_cats)

missing_cols <- setdiff(required_cols, names(df))
missing_cols


```

✅ Step 1.5 — Combine (pool) all years within state × race

```{r}
pooled <- df %>%
  group_by(state, race_ethnicity) %>%
  summarise(
    total_exits = sum(total_exits, na.rm = TRUE),
    dismissed = sum(dismissed, na.rm = TRUE),
    moved_out = sum(moved_out, na.rm = TRUE),
    not_determined = sum(not_determined, na.rm = TRUE),
    not_eligible = sum(not_eligible, na.rm = TRUE),
    part_b_eligible = sum(part_b_eligible, na.rm = TRUE),
    withdrawn = sum(withdrawn, na.rm = TRUE),
    .groups = "drop"
  )

dim(pooled)
head(pooled)

```

✅ Step 1.6 — Reshape to long by exit category (set up the 2×2 pieces)

```{r}
pooled_long <- pooled %>%
  tidyr::pivot_longer(
    cols = all_of(exit_cats),
    names_to = "category",
    values_to = "a"   # a = exits in this category for this race_ethnicity
  ) %>%
  mutate(
    b = total_exits - a,      # b = all other exits for this race_ethnicity
    flag_cat_gt_total = a > total_exits,
    flag_negative_b = b < 0
  )

dim(pooled_long)
head(pooled_long)

```

✅ Step 1.7 — Compute state totals for each category (the reference group building blocks)

```{r}
state_totals_long <- pooled %>%
  tidyr::pivot_longer(
    cols = all_of(exit_cats),
    names_to = "category",
    values_to = "state_event_total"
  ) %>%
  group_by(state, category) %>%
  summarise(
    state_total_exits = sum(total_exits, na.rm = TRUE),
    state_event_total = sum(state_event_total, na.rm = TRUE),
    .groups = "drop"
  )

dim(state_totals_long)
head(state_totals_long)

```

✅ Step 1.8 — Build the full 2×2 table (a, b, c, d) and compute odds ratios

```{r}
or_tbl <- pooled_long %>%
  left_join(state_totals_long, by = c("state", "category")) %>%
  mutate(
    # reference group = all other races in the same state
    c = state_event_total - a,
    d = (state_total_exits - total_exits) - c,

    # flags for data issues
    flag_zero_cell = (a == 0 | b == 0 | c == 0 | d == 0),
    flag_negative_cells = (c < 0 | d < 0),

    # Haldane–Anscombe correction if any cell is zero
    a_adj = if_else(flag_zero_cell, a + 0.5, as.numeric(a)),
    b_adj = if_else(flag_zero_cell, b + 0.5, as.numeric(b)),
    c_adj = if_else(flag_zero_cell, c + 0.5, as.numeric(c)),
    d_adj = if_else(flag_zero_cell, d + 0.5, as.numeric(d)),

    # odds ratio and log odds ratio
    or = (a_adj * d_adj) / (b_adj * c_adj),
    log_or = log(or)
  )

dim(or_tbl)
head(or_tbl)

```

✅ Step 1.9 — Select final columns and save for Shiny

```{r}
final_or_tbl <- or_tbl %>%
  select(
    state,
    race_ethnicity,
    category,
    a, b, c, d,
    or,
    log_or,
    flag_zero_cell,
    flag_negative_cells
  ) %>%
  arrange(state, category, race_ethnicity)

dim(final_or_tbl)
head(final_or_tbl)

```

✅ Step 1.11a — Add a “small cell” flag now (no CIs yet)

YAY I REMEMBERED TO FLAG CELLS SMALLER THAN 5 - MAKING LINA PROUD 🙌🏼

```{r}
final_or_tbl <- final_or_tbl %>%
  mutate(
    flag_small_cell_5 = (a < 5 | b < 5 | c < 5 | d < 5)
  )

final_or_tbl %>%
  summarise(
    n_rows = n(),
    n_small_cell_5 = sum(flag_small_cell_5, na.rm = TRUE),
    pct_small_cell_5 = mean(flag_small_cell_5, na.rm = TRUE)
  ) %>%
  print()

```

✅ **Option A — Add 95% confidence intervals for each odds ratio**

```{r}
final_or_tbl <- final_or_tbl %>%
  mutate(
    # recreate adjusted cells (same logic as before)
    a_adj = if_else(flag_zero_cell, a + 0.5, as.numeric(a)),
    b_adj = if_else(flag_zero_cell, b + 0.5, as.numeric(b)),
    c_adj = if_else(flag_zero_cell, c + 0.5, as.numeric(c)),
    d_adj = if_else(flag_zero_cell, d + 0.5, as.numeric(d)),

    # standard error of log(OR)
    se_log_or = sqrt(
      1 / a_adj +
      1 / b_adj +
      1 / c_adj +
      1 / d_adj
    ),

    # 95% confidence intervals
    ci_low  = exp(log_or - 1.96 * se_log_or),
    ci_high = exp(log_or + 1.96 * se_log_or)
  )

final_or_tbl %>%
  select(or, ci_low, ci_high, flag_small_cell_5) %>%
  head()

```

Final clean drop + reorder (run once)

```{r}
final_or_tbl <- final_or_tbl %>%
  select(
    state,
    race_ethnicity,
    category,
    a, b, c, d,
    or,
    log_or,
    se_log_or,
    ci_low,
    ci_high,
    flag_zero_cell,
    flag_small_cell_5
  ) %>%
  arrange(state, category, race_ethnicity)

names(final_or_tbl)

```

✅ Step — Save outputs (CSV first, then RDS)

```{r}
setwd("/Users/maiko/Desktop/capstone")
getwd()

```

```{r}
write.csv(
  final_or_tbl,
  "/Users/maiko/Desktop/capstone/data/analysis/state_avg_or_by_race_category_all_years.csv",
  row.names = FALSE
)

saveRDS(
  final_or_tbl,
  "/Users/maiko/Desktop/capstone/data/analysis/state_avg_or_by_race_category_all_years.rds"
)
```

```{r}

# Set working directory
setwd("/Users/maiko/Desktop/capstone/")

# 1. Import the CSV
years_combined_by_race_state <- read_csv("data/analysis/osep_exits_all_years_all_race_analysis.csv")

head(years_combined_by_race_state)
str(years_combined_by_race_state)
```

```{r}
# 2. Combine all years for each state/race combination
# Sum across years, keeping state and race_ethnicity separate

combined_by_state_race <- years_combined_by_race_state %>%
  group_by(state, race_ethnicity) %>%
  summarise(
    total_exits = sum(total_exits, na.rm = TRUE),
    dismissed = sum(dismissed, na.rm = TRUE),
    moved_out = sum(moved_out, na.rm = TRUE),
    not_determined = sum(not_determined, na.rm = TRUE),
    not_eligible = sum(not_eligible, na.rm = TRUE),
    part_b_eligible = sum(part_b_eligible, na.rm = TRUE),
    withdrawn = sum(withdrawn, na.rm = TRUE),
    .groups = 'drop'
  )

head(combined_by_state_race)
```

```{r}
# Look at South Dakota specifically
sd_data <- combined_by_state_race %>%
  filter(state == "South Dakota") %>%
  arrange(race_ethnicity)

sd_data
```

```{r}
# Look at South Dakota specifically
sd_data <- combined_by_state_race %>%
  filter(state == "South Dakota") %>%
  arrange(race_ethnicity) %>%
  mutate(
    dismissed_pct = round(100 * dismissed / total_exits, 2)
  ) %>%
  select(state, race_ethnicity, total_exits, dismissed, dismissed_pct, moved_out, not_determined, not_eligible, part_b_eligible, withdrawn)
```

```{r}
kable(sd_data %>% select(-state), 
      col.names = c("Race/Ethnicity", "Total Exits", "Dismissed for No Contact", "% Dismissed", "Moved Out", "Not Determined", "Not Eligible", "Part B Eligible", "Withdrawn"),
      caption = "South Dakota EI Exits by Race/Ethnicity (All Years Combined)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>%
  kable_classic(full_width = FALSE, html_font = "Times New Roman") %>%
  row_spec(0, extra_css = "line-height: 1;")
```

```{r}
setwd("/Users/maiko/Desktop/capstone/")

# 3. Save the combined dataset
saveRDS(combined_by_state_race, "data/analysis/combined_years_by_state_race.rds")
write_csv(combined_by_state_race, "data/analysis/combined_years_by_state_race.csv")
```
